---
title: "Time-varying Bradley-Terry model using Barycentric Interpolation"
author: "Erik Strumbelj, Blaz Krese"
output: pdf_document
---

## Model

The model is Bradley-Terry but team strength is allowed to vary over time. Latent team strength $\lambda$ is determined for each of $k$ teams at $m$ different nodes in time. The latent strengths for times between nodes ($\theta$) are interpolated using Barycentric rational interpolation. The $k$-th team is taken as the reference - its latent strength $\theta$ is assumed to be 0 always. The model also includes an additive term for home team advantage:

\begin{align*}
y_i |  \lambda, \Delta_\text{hta}, home, away, t &\sim \text{Bernoulli}\left(\frac{1}{1 + \exp(-(\theta_{home(i), t(i)} - \theta_{away(i), t(i)} + \Delta_\text{hta}))}\right) \\
\theta_{i,t} &=  \frac{\sum_{l = 1}^m (-1)^l \lambda_{i,l} / (t - t_l) }{\sum_{l = 1}^m (-1)^l / (t - t_l))}, \forall i \neq k \\
\theta_{k,\cdot} &= 0\\
\lambda_i &\sim N(0, 2), \forall i \neq k \\
\Delta_\text{hta} &\sim N(0, 1),
\end{align*}

where vectors $home(i)$/$away(i)$ and $t(i)$ are the indices of the home and away teams and the time of the $i$-th game, respectively. The prior on $\lambda$ is based on the competitive balance of the NBA - the worst team still has at least a 5-10% chance of beating the best team, which roughly coresponds to 2-3 difference in latent strength).

## Example: NBA basketball

```{r}
pct_train   <- 0.50
pct_test    <- 0.50 # the two 'pct' combined can't exceed 1.0
n_teams     <- 3    # between 2 and 30
nodes       <- 5
iter_sample <- 1000
iter_warmup <- 200
n_chains    <- 1
seed        <- 1

df <- readRDS("NBA_dataset.rds")
print(head(df))
```

The dataset consists of `r nrow(df)` regular season NBA games between `r min(df$Date)` and  `r max(df$Date)`. The columns include the outcome (y), an indexing of the teams (Object), the relative time in days (Time), and home and away win probabilities (Prob) derived from bookmaker odds.

We'll take only the first `r n_teams` teams use `r pct_test` of the observations for training and `r pct_test` for testing. The number of interpolation nodes is set to `r nodes`. The computation time will increase substantially if we use all 30 teams and in particular if we increase the number of nodes.


```{r}
library(rstan)

# Train/Test split
df      <- df[df$Object_a <= n_teams & df$Object_b <= n_teams,]
n       <- nrow(df)
n_train <- floor(pct_train * n)
n_test  <- floor(pct_test  * n)
idx     <- sample(1:nrow(df), n_train + n_test, rep = F)
idx_te  <- idx[1:n_test]
idx_tr  <- idx[-c(1:n_test)]

x_train <- df[idx_tr, colnames(df) != "y"]
y_train <- df[idx_tr, colnames(df) == "y"]
x_test  <- df[idx_te, colnames(df) != "y"]
y_test  <- df[idx_te, colnames(df) == "y"]

# Derived data
objs      <- sort(unique(unlist(x_train[c("Object_a", "Object_b")])))
tt        <- sort(unique(rbind(x_train, x_test)$Time))
tt_train  <- sort(unique(x_train$Time))
idx_train <- sapply(x_train$Time, function(x) which(x == tt_train)) 
idx_test  <- sapply(x_test$Time, function(x) which(x == tt))
t_k       <- seq(min(tt_train), max(tt_train), 
                 (max(tt_train) - min(tt_train)) / (nodes + 1)) + pi/10
t_k       <- t_k[c(-1, -length(t_k))]
weights  <- (-1)^(1:length(t_k))

# Prepare data for Stan
stan_data <- list(N_teams = length(objs), n_nodes = nodes,
                  t_k = as.array(t_k), w = as.array(weights),
                  N_tt = length(tt), tt = tt, N_tt_train = length(tt_train),
                  tt_train = tt_train, N_train = nrow(x_train),
                  N_test = nrow(x_test),
                  i_train = cbind(x_train$Object_a, x_train$Object_b),
                  i_test = cbind(x_test$Object_a, x_test$Object_b),
                  idx_train = as.array(idx_train),
                  idx_test = as.array(idx_test), y = y_train)

# Compile and sample
sm  <- stan_model(file = "barycentric.stan")
res <- sampling(sm, stan_data, iter = iter_sample,
                warmup = iter_warmup, chains = n_chains, seed = seed,
                control = list(max_treedepth = 20, adapt_delta = 0.8))

print(res, pars = c("hta"))

samples <- extract(res)
```


```{r}
# visualize over time
library(ggplot2)

th_out <- samples$th
p      <- colMeans(samples$p)
hta    <- samples$hta
dates  <- as.Date(df$Date[match(tt, df$Time)])
x <- NULL
for (i in seq(length(objs))) {
  tmp <- t(th_out[,i,])   # No ordering, because tt is sorted
  x <- rbind(x, data.frame(Team = df$Abbr.x[match(i, df$Object_a)], 
                           Time = dates, 
                           Theta = rowMeans(tmp)))
}

ggplot(x, aes(x = Time, y = Theta, group = Team, colour = Team)) + 
  geom_line() + ylab("latent strength (posterior mean)") +
  xlab("time")
```

## Model code

```{r comment=''}
cat(readLines('barycentric.stan'), sep = '\n')
```

